---
title: "p8105_hw2_zn2220"
author: "Ziang Niu"
date: "2025-09-24"
output: github_document
---

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
library(haven)
```

# Problem 1

## 1.1

### First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_df <- read_csv("data/pols-month.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(
    mon, into = c("year", "month", "day"), sep = "-"
  ) %>% 
  mutate(
    year = as.integer(year),
    month = factor(
      month.name[as.integer(month)],
      levels = month.name,
      ordered = TRUE
    )
  ) %>% 
  mutate(
    president = case_when(
      prez_dem == 1 ~ "dem",
      prez_gop == 1 ~ "gop",
      TRUE ~ NA_character_
    )
  ) %>% 
  select(-prez_dem, -prez_gop, -day) %>% 
  select(year, month, everything())

```

## 1.2

### Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp_df = 
  read_csv("data/snp.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(
    date, into = c("month", "day", "year"), sep = "/"
  ) %>% 
  mutate(
    year = if_else(
      as.integer(year) <= 15,
      as.integer(year) + 2000,
      as.integer(year) + 1900
    ),
    month = factor(
      month.name[as.integer(month)], 
      levels = month.name, ordered = TRUE
    )
  ) %>% 
  select(-day) %>% 
  arrange(year, month) %>% 
  select(year, month, everything())

```

## 1.3

### Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
une_df = 
  read_csv("data/unemployment.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment",
  ) %>% 
  mutate(
    month = factor(
      month,
      levels = tolower(month.abb),
      labels = month.name,
      ordered = TRUE 
    )
  ) %>% 
  select(year, month, everything())
```

## 1.4

### Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
overall_df <- pols_df %>% 
  full_join(snp_df, by = c("year", "month")) %>% 
  full_join(une_df, by = c("year", "month")) %>% 
  janitor::clean_names()

```

## 1.5

### Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

The pols_df contains:

```{r}
summary(pols_df)
```

The snp_df contains:

```{r}
summary(snp_df)
```

The une_df contains:

```{r}
summary(une_df)
```

The combined dataset `overall_df` can be summarized with the following R functions:

- The dimension is `r paste(dim(overall_df), collapse = " x ")`.  
- The range of year is from `r min(overall_df$year, na.rm = TRUE)` to `r max(overall_df$year, na.rm = TRUE)`.  
- The key variables are: `r paste(names(overall_df), collapse = ", ")`.  
- A descriptive summary of each variable is:

```{r}
summary(overall_df)
```

### Note

Create the data key.

```{r}
overall_df <- overall_df %>%
  mutate(date = lubridate::make_date(year, as.integer(month)))
```

# Problem 2

## 2.1

### Read and clean the Mr. Trash Wheel sheet:specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel. Use reasonable variable names. Omit rows that do not include dumpster-specific data. Round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
mtw_df <- read_excel(
  "data/202409 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  skip = 1
) %>%
  select(-...15, -...16) %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year),
    wheel = "Mr. Trash Wheel"
  )
```

## 2.2

### Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}
ptw_df <- read_excel(
  "data/202409 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
  skip = 1
) %>%
  select(-starts_with("...")) %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    year = as.integer(year),
    wheel = "Professor Trash Wheel"
  )

gtw_df <- read_excel(
  "data/202409 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda Trash Wheel",
  skip = 1
) %>%
  select(-starts_with("...")) %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    year = as.integer(year),
    wheel = "Gwynnda Trash Wheel"
  )
```

Combine.
```{r}
wheelall_df <- bind_rows(mtw_df, ptw_df, gtw_df) %>%
  filter(!is.na(date)) %>%
  arrange(date, wheel)
```

## 2.3

### Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?

The combined dataset of Trash Wheel collections contains `r nrow(wheelall_df)` observations, with key variables including:

```{r}
names(wheelall_df)
```

Professor Trash Wheel collected a total of `r sum(wheelall_df$weight_tons[wheelall_df$wheel == "Professor Trash Wheel"])` tons of trash. In addition, Gwynnda Trash Wheel collected `r sum(wheelall_df$cigarette_butts[wheelall_df$wheel == "Gwynnda Trash Wheel" & wheelall_df$year == 2022 & wheelall_df$month == "June"])` cigarette butts in June of 2022.

# Problem 3

## 3.1

### Create a single, well-organized dataset with all the information contained in these data files. To that end: import, clean, tidy, and otherwise wrangle each of these datasets; check for completeness and correctness across datasets (e.g. by viewing individual datasets and monitoring warning messages); merge to create a single, final dataset; and organize this so that variables and observations are in meaningful orders.

Clean and see.

```{r}
zc_df <- read_csv("data/Zip Codes.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names()

nyc_df <- read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names()

glimpse(zc_df)
glimpse(nyc_df)
```

Transformation

```{r}
length(unique(zc_df$zip_code))
length(unique(nyc_df$region_name))

zc_df <- zc_df %>%
  mutate(zip_code = as.character(zip_code))

nyc_df <- nyc_df %>%
  rename(zip_code = region_name) %>%
  mutate(zip_code = as.character(zip_code))

common_zips <- intersect(zc_df$zip_code, nyc_df$zip_code)
length(common_zips)

setdiff(nyc_df$zip_code, zc_df$zip_code)

```

Combination.

```{r}
tidy_df <- nyc_df %>%
  left_join(zc_df, by = "zip_code")

tidy_df <- tidy_df %>%
  relocate(zip_code, city, county, state, .before = 1)

glimpse(tidy_df)
summary(tidy_df)

```

## 3.2

### Briefly describe the resulting tidy dataset. How many total observations exist? How many unique ZIP codes are included, and how many unique neighborhoods?

The resulting tidy dataset has `r nrow(tidy_df)` total observations. It includes `r n_distinct(tidy_df$zip_code)` unique ZIP codes and `r n_distinct(tidy_df$neighborhood)` unique neighborhoods.

## 3.3

### Which ZIP codes appear in the ZIP code dataset but not in the Zillow Rental Price dataset? Using a few illustrative examples discuss why these ZIP codes might be excluded from the Zillow dataset.

```{r}
zips_only_in_ref <- setdiff(zc_df$zip_code, nyc_df$zip_code)

head(zips_only_in_ref, 10)
```

Some ZIP codes appear in the reference file but not in Zillow’s rental dataset because they do not represent active residential markets. In many cases these areas are primarily commercial or industrial, or they contain very few rental properties, making it difficult for Zillow to calculate a reliable rent index. In other instances, the ZIP codes may be newly created or used mainly for postal delivery rather than for households. Finally, the scope of the Zillow dataset is limited to neighborhoods where sufficient rental activity is observed, so ZIP codes outside the core residential areas of New York City are naturally excluded.

## 3.4

### Rental prices fluctuated dramatically during the COVID-19 pandemic. For all available ZIP codes, compare rental prices in January 2021 to prices in January 2020. Make a table that shows the 10 ZIP codes (along with the borough and neighborhood) with largest drop in price from January 2020 to 2021. Comment.

```{r}
result <- tidy_df %>%
  mutate(diff = `x2021_01_31` - `x2020_01_31`) %>%
  arrange(diff) %>%
  slice_head(n = 10) %>%
  select(zip_code, county, neighborhood, diff)
```

The values range from about -912.6 to -684.9, and Lower Manhattan appears most often with the lowest values.
























