---
title: "p8105_hw2_zn2220"
author: "Ziang Niu"
date: "2025-09-24"
output: github_document
---

```{r}
library(tidyverse)
```

# Problem 1

## 1.1

### First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_df <- read_csv("data/pols-month.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(
    mon, into = c("year", "month", "day"), sep = "-"
  ) %>% 
  mutate(
    year = as.integer(year),
    month = factor(
      month.name[as.integer(month)],
      levels = month.name,
      ordered = TRUE
    )
  ) %>% 
  mutate(
    president = case_when(
      prez_dem == 1 ~ "dem",
      prez_gop == 1 ~ "gop",
      TRUE ~ NA_character_
    )
  ) %>% 
  select(-prez_dem, -prez_gop, -day)

```

## 1.2

### Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp_df = 
  read_csv("data/snp.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(
    date, into = c("month", "day", "year"), sep = "/"
  ) %>% 
  mutate(
    year = if_else(
      as.integer(year) <= 15,
      as.integer(year) + 2000,
      as.integer(year) + 1900
    ),
    month = factor(
      month.name[as.integer(month)], 
      levels = month.name, ordered = TRUE
    )
  ) %>% 
  select(-day) %>% 
  arrange(year, month) %>% 
  select(year, month, everything())

```

## 1.3

### Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
une_df = 
  read_csv("data/unemployment.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment",
  ) %>% 
  mutate(
    month = factor(
      month,
      levels = tolower(month.abb),
      labels = month.name,
      ordered = TRUE 
    )
  )
```

## 1.4

### Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
overall_df <- pols_df %>%
  full_join(snp_df, by = c("year", "month")) %>%
  full_join(une_df, by = c("year", "month")) %>%
  janitor::clean_names()

```

## 1.5

### Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

The pols_df contains:

```{r}
summary(pols_df)
```

The snp_df contains:

```{r}
summary(snp_df)
```

The une_df contains:

```{r}
summary(une_df)
```

The combined dataset `overall_df` can be summarized with the following R functions:

- The dimension is `r paste(dim(overall_df), collapse = " x ")`.  
- The range of year is from `r min(overall_df$year, na.rm = TRUE)` to `r max(overall_df$year, na.rm = TRUE)`.  
- The key variables are: `r paste(names(overall_df), collapse = ", ")`.  
- A descriptive summary of each variable is:

```{r}
summary(overall_df)
```



